Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time

### 2025-02-05 ms-3/ms.tex

\begin{abstract}
Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
However, measures of temporal volatility in incidence, and how volatility changes over time, are often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect rapid shifts in the dispersion parameter, which may signal new epidemic phases. 
We apply the method to COVID-19 incidence data in 144 United States (US) counties from the January 1st, 2020 to March 23rd, 2023.
Theory predicts that dispersion should be inversely proportional to incidence, however our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties. 
In particular, dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
These findings suggest that heterogeneity in transmission, susceptibility, and reporting could play important roles in driving large surges and extending epidemic duration. 
The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmission, with potential applications as leading indicators for public health response.
\end{abstract}

\clearpage
\section*{Author summary}
Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public Traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic of incidence time series. 
Quantifying variability around incidence trends can elucidate important underlying processes, including transmission heterogeneity.
We developed a statistical framework to quantify temporal changes in case count dispersion within a single time series and applied the method to COVID-19 case count data. 
We found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone. 
Dispersion increased around peaks in incidence such as the major surge in cases in 2022, and dispersion also increased as the pandemic progressed. 
These increases potentially indicate transmission heterogeneity, changes in the susceptibility landscape, or that there were changes in reporting.
Shifts in dispersion can also indicate shifts in epidemic phase, so our method provides a way for public health officials to anticipate and manage changes in epidemic regime and the drivers of transmission.

\clearpage
\section*{Introduction}
Time series of infectious disease incidence appear, to varying degrees, ``noisy'', showing higher frequency fluctuations (e.g., day-to-day or week-to-week fluctations) around trends at the broader temporal ranges typical for epidemic curves (e.g., months or years).
Short-term fluctuations in incidence time series are caused in part by variable reporting, but may also reflect the population-level impacts of transmission heterogeneity, and changes in the landscape of susceptiblility~\citep{lloyd2005superspreading,lloyd2007maximum, lau2017spatial, dalziel2018urbanization, kirkegaard2021superspreading, sun2021transmission,guo2023statistical,ko2023time}.
Metrics of variability in incidence time series may therefore carry information regarding underlying drivers of transmission, and offer a relatively unexplored avenue for understanding epidemic dynamics.

@@

An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see Supplemental Information). 
For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
In a simple example relevant to surge capacity in healthcare systems, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.

In a sufficiently large host population, and when the infectious pathogen can be assumed to spread in nonoverlapping generations, the number of infections each generation is often modeled as
\begin{equation}
    I_{t+1} \sim NB(\mu_t = R_t I_t, \theta_t = I_t)
\end{equation}
where time-varying reproductive number $R_t$ gives the expected number of secondary infections aqcuired from an infected indivual at time $t$, and the generation time is set to 1 without loss of generality~\citep{kendall1949stochastic, bjornstad2002dynamics}. 
Setting $\theta_t = I_t$ arises from the assumption that individuals who acquire the infection at time $t$ form independent lineages with identically distributed local rate parameters.
In applications this model for $theta$ beomes $\theta_t = C_t / \rho_t$ where $C_t$ represents reported cases and $\rho$ the reporting rate, which relates reported cases to the true number of infections as $C_t = \rho_t I_t$.
However, this requires that susceptible depletion in one lineage does not affect another, that transmission rates are equal across lineages, and that reporting rates do not vary across lineages.

In practice, these assumption will not often hold, and our aim in this paper is to develop, test and apply an alternative approach, which makes data-driven estimates of $\theta_t$, including identifying timepoints when $\theta$ is changing rapidly, which may help to reveal the impacts of heterogeneity in transmission, susceptibility, and reporting.

\section*{Methods}
By definition incidence volatility is fast relative to broadscale epidemic dynamics. 
Consequently, in order to estimate incidence volatility we first model incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review}. To allow for diverse shapes in the broadscale epidemic dynamics, spline modeling is conducted in a moving window such that for each half of the window
\begin{equation}
  \log\left(\dfrac{\mu_t}{N}\right)  = \sum_{j=1}^J \beta_j^{(t)} h_j(t)
\end{equation}
where $N$ represents population size, $h_j(t)$ are basis functions, the degrees of freedom is equal to the number of knots $k$ for the natural spline, $J = k + d +1$, where $d$ is the degree of the polynomial, and $\beta_j^{(t)}$ are fitted parameters. The window has half-width $\Delta$, centered at $t$, i.e., extending from $t-\Delta$ to $t + \Delta$. 
The degrees of freedom (number of knots) to be used for the natural splines, and the width of the moving window will depend on the application.
Explaination of the specific choices we used for our application to COVID-19 cases in US counties is described below.

>>>>

\begin{Abstract}
Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
However, measures of temporal volatility in incidence, and how volatility changes over time, are often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect rapid shifts in the dispersion parameter, which may signal new epidemic phases. 
We apply the method to COVID-19 incidence data in 144 United States (US) counties from the January 1st, 2020 to March 23rd, 2023.
Theory predicts that dispersion should be inversely proportional to incidence, however our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties. 
In particular, dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
These findings suggest that heterogeneity in transmission, susceptibility, and reporting could play important roles in driving large surges and extending epidemic duration. 
The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmission, with potential applications as leading indicators for public health response.
\end{abstract}

\clearpage
\section*{Author summary}
Understanding patterns in infectious disease incidence is crucial for understanding epidemic dynamics and for developing effective public health policy.
Traditional metrics used to quantify incidence patterns often overlook variability as an important characteristic of incidence time series. 
Quantifying variability around incidence trends can elucidate important underlying processes, including transmission heterogeneity.
We developed a statistical framework to quantify temporal changes in case count dispersion within a single time series and applied the method to COVID-19 case count data. 
We found that conspicuous shifts in dispersion occurred across counties concurrently, and that these shifts were not explained by incidence alone. 
Dispersion increased around peaks in incidence such as the major surge in cases in 2022, and dispersion also increased as the pandemic progressed. 
These increases potentially indicate transmission heterogeneity, changes in the susceptibility landscape, or that there were changes in reporting.
Shifts in dispersion can also indicate shifts in epidemic phase, so our method provides a way for public health officials to anticipate and manage changes in epidemic regime and the drivers of transmission.

\clearpage
\section*{Introduction}
Time series of infectious disease incidence appear, to varying degrees, ``noisy'', showing higher frequency fluctuations (e.g., day-to-day or week-to-week fluctations) around trends at the broader temporal ranges typical for epidemic curves (e.g., months or years).
Short-term fluctuations in incidence time series are caused in part by variable reporting, but may also reflect the population-level impacts of transmission heterogeneity and changes in the landscape of susceptiblility~\citep{lloyd2005superspreading,lloyd2007maximum, lau2017spatial, dalziel2018urbanization, kirkegaard2021superspreading, sun2021transmission,guo2023statistical,ko2023time}.
Metrics of variability in incidence time series may therefore carry information regarding underlying drivers of transmission, and offer a relatively unexplored avenue for understanding epidemic dynamics.

@@

An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see derivation in S1 Text). 
For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
In a simple example relevant to surge capacity in healthcare systems, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) as expected for a Poisson process with the same incidence rate.

In a sufficiently large host population, and when the infectious pathogen can be assumed to spread in nonoverlapping generations, the number of infections each generation is often modeled as
\begin{equation}
    I_{t+1} \sim NB(\mu_t = R_t I_t, \theta_t = I_t)
\end{equation}
where time-varying reproductive number $R_t$ gives the expected number of secondary infections aqcuired from an infected indivual at time $t$, and the generation time is set to 1 without loss of generality~\citep{kendall1949stochastic, bjornstad2002dynamics}. 
Setting $\theta_t = I_t$ arises from the assumption that individuals who acquire the infection at time $t$ form independent lineages with identically distributed local rate parameters.
In applications, this model for $theta$ beomes $\theta_t = C_t / \rho_t$ where $C_t$ represents reported cases and $\rho$ the reporting rate, which relates reported cases to the true number of infections as $C_t = \rho_t I_t$.
However, this requires that susceptible depletion in one lineage does not affect another, that transmission rates are equal across lineages, and that reporting rates do not vary across lineages.

In practice, these assumption will not often hold, and our aim in this paper is to develop, test and apply an alternative approach which produces data-driven estimates of $\theta_t$, including identifying timepoints when $\theta$ is changing rapidly, which may help to reveal the impacts of heterogeneity in transmission, susceptibility, and reporting.

\section*{Methods}
By definition incidence volatility is fast relative to broadscale epidemic dynamics.

Consequently, in order to estimate incidence volatility we first modeled incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review}. To allow for diverse shapes in the broadscale epidemic dynamics, spline modeling was conducted within a moving window such that for each half of the window
\begin{equation}
  \log\left(\dfrac{\mu_t}{N}\right)  = \sum_{j=1}^J \beta_j^{(t)} h_j(t)
\end{equation}
where $N$ represents population size, $h_j(t)$ are basis functions, the degrees of freedom is equal to the number of knots $k$ for the natural spline,  $J = k + d +1$, where $d$ is the degree of the polynomial, and $\beta_j^{(t)}$ are fitted parameters. The window has half-width $\Delta$, centered at $t$, i.e., extending from $t-\Delta$ to $t + \Delta$. 
The degrees of freedom (number of knots) to be used for the splines, and the width of the moving window will depend on the application.
Explanation of the specific choices we used for our application to COVID-19 cases in US counties is described below.
